### 2013-11-06 src2/thesis.tex
update

% ---------------------------------------------------------------------------
% Tryckortssida
% ---------------------------------------------------------------------------

@@

{\noindent\large\bf\thesistitle} \\
\noindent Bo Joel Svensson \\
\noindent ISBN \\

@@

\noindent Technical Report 101D\\
\noindent ISSN                 \\
\noindent \dept \\
\noindent \group\\

@@

Graphics Processors (GPUs) trade programmer convenience for increased parallelism. 
For example, where CPUs dedicate large portions of chip area to caches, branch 
prediction and Instruction level parallelism (ILP), GPUs put additional cores. 
This comes at a price of increased programmer effort. GPUs have programmer managed 
shared memories, conditionals are troublesome from a performance 
perspective and memory access patterns can make or break the performance of an 
application. With Obsidian we try to find the right level of abstraction for 
programming GPUs. We try to strike a balance between low-level control of details 
that influence performance and high-level abstractions to increase productivity and 
make programs more terse. The end goal with Obsidian is a language for writing 
high performance GPU programs that compare well to code written by CUDA 
(a C-dialect for GPU programming) experts, provided that both programmers have 
similar GPU expertise. The Obsidian programmer would however do so in a more 
composable manner. Obsidian programs are compiled into CUDA.

Obsidian is geared towards the programmer who is interested in GPU hardware. Making 
good use of Obsidian requires that the programmer is willing to make an effort 
of learning about GPUs and how to program them efficiently. However, another case 
is a programmer who does not want to invest that effort but wants to write a high-level 
parallel program and have it run reasonably well on different parallel architectures. 
In this case a language with even higher level of abstraction is preferable. It then 
becomes the job of a more advanced compiler to turn that into code for the targetted 
platforms. The EmbArBB language targets this programmer. EmbArBB exposes a set of high-level 
operations on arrays, such as {\tt map}, {\tt reduce} and {\tt scan}. The programmer 
has no power over the parallel strategy that should be used to implement these 
operations but instead rely on a JIT-compiler (Just In Time-compiler). EmbArBB is a 
Haskell embedding of Intel ArBB that was released as an embedded language in C++.

@@

\node[anchor=west] (acc) at (1,2.7) {Accelerate\\Nikola\\Thrust};
\node[anchor=west] (acc) at (1,1.7) {CUB};
\node[anchor=west] (repa) at (0,2.6) {DPH\\Nesl\\Repa\\Meta-repa};
\node[anchor=west] (mixed) at (0,1.7) {Cilk+\\ TBB\\ Feldspar};
\node[anchor=west] (accelerator) at (2,1.7) {Accelerator};

>>>>

% ---------------------------------------------------------------------------
% Tryckortssida
% ---------------------------------------------------------------------------
%ditt ISBN är:  978-91-7385-939-4
%ditt Ny serie nr:  3620
%ISSN för publ.serien Doktorsavhandlingar vid Chalmers tekniska högskola : 0346-718X

@@

{\noindent\large\bf\thesistitle} \\
\noindent Bo Joel Svensson \\
\noindent ISBN 978-91-7385-939-4 \\

@@

\noindent Doktorsavhandlingar vid Chalmers tekniska högskola \\
\noindent Ny serie nr 3620 \\
\noindent ISSN 0346-718X \\

\noindent Technical Report 101D\\
\noindent \dept \\
\noindent \group\\

@@

Graphics Processors (GPUs) trade programmer convenience for increased parallelism. 
For example, where CPUs dedicate large portions of chip area to caches, branch 
prediction and Instruction level parallelism (ILP), GPUs put additional cores. 
This comes at a price of increased programmer effort. GPUs have programmer managed 
shared memories, conditionals are troublesome from a performance 
perspective and memory access patterns can make or break the performance of an 
application. With Obsidian we try to find the right level of abstraction for 
programming GPUs. We try to strike a balance between low-level control of details 
that influence performance and high-level abstractions to increase productivity and 
make programs more terse. The end goal with Obsidian is a language for writing 
high performance GPU programs that compare well to code written by CUDA~\citet{CUDA}  
(a C-dialect for GPU programming) experts, provided that both programmers have 
similar GPU expertise. The Obsidian programmer would however do so in a more 
composable manner. Obsidian programs are compiled into CUDA.

Obsidian is geared towards the programmer who is interested in GPU hardware. Making 
good use of Obsidian requires that the programmer is willing to make an effort 
of learning about GPUs and how to program them efficiently. However, another case 
is a programmer who does not want to invest that effort but wants to write a high-level 
parallel program and have it run reasonably well on different parallel architectures. 
In this case a language with even higher level of abstraction is preferable. It then 
becomes the job of a more advanced compiler to turn that into code for the targetted 
platforms. The EmbArBB language targets this programmer. EmbArBB exposes a set of high-level 
operations on arrays, such as {\tt map}, {\tt reduce} and {\tt scan}. The programmer 
has no power over the parallel strategy that should be used to implement these 
operations but instead rely on a JIT-compiler (Just In Time-compiler). EmbArBB is a 
Haskell embedding of Intel ArBB that was released as an embedded language in C++. 
EmbArBB relies entirely on the ArBB system for code generation and optimisation.

\section {Obsidian: An embedded language for GPU kernel implementation}

In 2006 NVIDIA released their first Compute Unified Device Architecture (CUDA) 
GPU. With CUDA, NVIDIA also made it a lot easier for programmers to use GPUs
for general purpose programming. A dialect of C, called CUDA C, was released 
specifically to simplify the implementation of general purpose computations on GPUs.

Before CUDA, programmers still tried (very ingeniously) to make use of GPUs for 
non-graphics applications via the graphics API. The computation needed to be expressed 
in graphics terms and the result computed through an act of rendering. Of course, since the 
available abstractions targetted at the graphics domain, this added extra
burden  on the programmer. CUDA was NVIDIA's first response to this problem. With CUDA 
there is a dialect of C, extended with GPU specific functionality.

Having a language specifically made for general purpose programming on the GPU, such as 
CUDA, is a big step up from earlier. But CUDA still offers a very low-level interface to 
programming the GPU, which is preferred in a situation where the programmer needs to 
think about hardware specific details to get the best performance. This CUDA provides, 
but at the price that it is hard to build upon, abstract and reuse prior programming 
effort. This is where we are trying to go one step further with Obsidian.

\subsection{GPU hardware}

Before going into programming of GPUs, it is important to have some background. The 
GPU programming model exposed by CUDA very much mirrors the underlying hardware. Some 
of the details that make GPU programming hard is more apparent when looking at the 
underlying hardware.

A CUDA GPU is built around a single kind of processor (as opposed to the different 
kinds of processors found in earlier GPUs). The processors in the GPU (called MPs, 
MultiProcessors) all contain a number of cores called SPs (Streaming Processors). 
Each MP also contains local memory, called shared memory since it can be accessed 
by all of the SPs in that MP. The number of MPs varies over the available GPUs; cheaper 
GPUs have as few as one MP, and as you go up in price the number of MPs increases.

\subsection{Writing efficient GPU Kernels} 
 \emph{now takes up both CUDA and Obsidian aspects at the same time, pref. with code examples}

@@

\node[anchor=west] (acc) at (1,2.7) {Accelerate\\Nikola\\Thrust};
\node[anchor=west] (acc) at (1,1.7) {CUB};
\node[anchor=west] (repa) at (0,2.5) {DPH\\Nesl\\Repa\\Meta-repa\\Feldspar};
\node[anchor=west] (mixed) at (0,1.7) {Cilk+\\ TBB};
\node[anchor=west] (accelerator) at (2,1.7) {Accelerator};
